Elastic Scattering in General Relativistic Ray Tracing for Neutrinos

### 2017-07-25 paper.tex
Remove troublesome \todo macro
It appears my use of the \todo macro introduces a lot of compile warnings
"Underfull \hbox (badness 10000) ...". Till I fix that or rearrange my
\todo placements, I'll just turn them off.
Expand analysis of Post-Bounce test

\newcommand{\todo}[1]{\marginpar{\tiny{\textcolor{red}{#1}}}}
%\renewcommand\todo[1]{} % if you wish to hide all todo notes

@@

\subsection{Post-Bounce Collapse Profile:
  Testing Scattering}
\label{ssec:test_collapse}
To test our scattering treatments we calculate neutrino fields outside a
collapsed 15~${\rm M}_\odot$ star, 100~ms after core bounce,
comparing ray tracing fields to those from a Monte Carlo transport calculation.
Elastic scattering interior to the shock at $r\approx150$~km
significantly modifies the neutrinos' spectra,
and the extended envelope outside the shock becomes a source of
higher-than-average-energy neutrinos.

@@

The matter profile and background field are stored on a spherical
pseudo-spectral grid composed of 11 spherical-shell subdomains
\todo{cite early SpEC paper}
comprising a total of 62 radial grid points spaced approximately
logarithmically across $r\in(0,740)$~km.
The background scattering field $\Phi(\varepsilon)$ is supplied by
\lstinline{GR1D} in the form of $J(\varepsilon_i)$,
with $\varepsilon_i$ representing 18 energy groups
identical to those in the \lstinline{NuLib} table described in
Sec.~\ref{ssec:test_equilibrium}.
For the \emph{spectral} method we use $J(\varepsilon)$ directly,
using zeroth-order interpolation between energy groups;
for the \emph{gray} method we use $J=\sum J(\varepsilon_i) \Delta\varepsilon_i$,
and $G=\sum J(\varepsilon_i) \Delta\varepsilon_i/\varepsilon_i$ instead.

@@

As expected the average energies from both
the scattering envelope ($\cos A \lesssim 0.95$) and
the bright core ($\cos A\gtrsim 0.995 $)
are badly characterized by the \emph{noscat} treatment.
In this case for $\nu_x$ the major effect of elastic scattering
is to decrease the average energy of neutrinos coming from the core
and increase the average energy of neutrinos coming from the envelope.

@@

We can make some quantitative sense of the scattered energies in
Fig.~\ref{fig:avg_eps_collapse} from the formal solution explored in the
simple configurations above.
In particular, the average energy in the scattering envelope
($\cos A\lesssim 0.95$) is related to the spectrum in the direction
of the interior ($\cos A \sim 1$).
The relation is quite simple if we simplify our realistic model
to that of a homogeneous matter profile and
background scattering field.

In the envelope the local temperature is low
and the ratio $\chi^*_a/\chi_s$ small,
so $f_{\rm SE}$ strongly dominates over $f_{\rm AE}$.
By expanding the exponential of
Eqn.~\ref{eqn:fse_homogeneous_general} as we did for
Eqn.~\ref{eqn:fae_homogeneous_limits},
and furthermore factoring out the dominant ${\rm energy}^2$ dependence from the
scattering opacity $\chi_s\equiv\varepsilon^2 \zeta_s$, we have
\begin{equation}
  \label{eqn:fse_homogeneous_thin_limit}
  f_{\rm SE}(\varepsilon)
  = \varepsilon^2 \zeta_s \,\Phi(\varepsilon)\,s,
\end{equation}
where $s$ is the length of scattering envelope passed through by the ray.
The average energy of the scattered field is
\begin{equation}
  \label{eqn:avg_eps_sc}
  \langle \varepsilon \rangle^{\rm sc}
  = \frac{\int d\varepsilon\,\varepsilon^3 f}{\int d\varepsilon\,\varepsilon^2 f}
  = \frac{\int d\varepsilon\,\varepsilon^5 \Phi}{\int d\varepsilon\,\varepsilon^4 \Phi}.
\end{equation}

The spectrum of the background field $\Phi(\varepsilon)$ is well-approximated
as the \emph{scat} solution for $\cos A\sim 1$, since that is the dominant
source direction for neutrinos.
And because the scattering envelope is optically thin at all energies,
we can assume the neutrino spectrum is essentially unchanged in its passage
through the envelope.
In order to estimate $\langle \varepsilon \rangle^{\rm sc}$ analytically, we may
approximate the background field $\Phi(\varepsilon)$ analytically.
Direct Fermi-Dirac fits using a temperature and chemical potential
representative of a physical neutrinosurface fair poorly since neutrinos of
different energies decouple from the matter at different radii, over which
the thermodynamic state varies substantially.
Phenomenological Fermi-Dirac fits work well;
but so do pinched spectral fits which are much simpler \cite{miri????}:
\begin{equation}
  \label{eqn:pinched_spectra}
  \Phi^{\rm pi}(\varepsilon) \propto
  \varepsilon^{\alpha-2}
  \exp\left(-(\alpha+1)\frac{\varepsilon}{\langle\varepsilon\rangle}\right),
\end{equation}
(where our definition differs from that of \cite{miri????} by a factor of
$\varepsilon^2$, since we define our distribution function to be dimensionless
in natural units $\{\hbar,c\}=1$).

Pinching parameters (calculated by eye) for the species
$\{\nu_e,\bar{\nu}_e,\nu_x\}$ are $\alpha\approx\{3.6,5.1,2.3\}$,
when we use the average energies from the \emph{spectral}
\emph{scat} method in Tab.~\ref{tab:nsns_rt_vs_m1}.
Energy moments of pinched spectra like those in Eqn.~\ref{eqn:avg_eps_sc}
have a simple analytic form so that Eqn.~\ref{eqn:avg_eps_sc} becomes
\begin{align}
  \label{eqn:pinched_avg_eps_sc_1}
  \langle\varepsilon\rangle^{\rm sc}
  &\approx \frac{\alpha+3}{\alpha+1} \langle\varepsilon\rangle, \\
  %\label{eqn:pinched_avg_eps_sc_2}
  %&= \{1.4,1.3,1.6\} \langle\varepsilon\rangle, \\
  \label{eqn:pinched_avg_eps_sc_3}
  &\approx \{16,18,28\}\,~{\rm MeV},
\end{align}
again for $\{\nu_e,\bar{\nu}_e,\nu_x\}$ respectively.
These analytic predictions agree with the average energies of the scattering
envelope to within 20\%
% From by-eye examinations of <eps>(cosA) plot using data from
% MCDATA=$DIR/ResultsSherwood-02
% RTDATA_GR=$DIR/ResultsBrett-B-03/Test-scat-01/
% RTDATA_SP=$DIR/ResultsBrett-A-03/Test-scat-01/
%          nue     nua     nux
of the Monte Carlo and the \emph{scat} ray tracing calculations,
even though we have made some drastic simplifications to our model.

@@

However moment evolutions with multiple energy groups are still rare.
Most simulations employ a gray moment scheme,
evolving only the energy-integrated moments, $J$ and $H^\alpha$
and sometimes the number density $G$.
If such is the case we resort to the \emph{gray} method,
by reconstructing the energy-dependent source terms
from gray moments, assuming a diluted Fermi-Dirac spectrum.
We use the following procedure:
\begin{enumerate}
\item
  Interpolate the fluid temperature, $T$,
  equilibrium neutrino chemical potential, $\eta_\nu$,
  and evolved neutrino energy and number densities in the fluid frame,
  $J$ and $G$, from the simulation grid.
\item
  Compute the average neutrino energy in the fluid frame
  \begin{equation}
    \langle \varepsilon \rangle \equiv J/G.
  \end{equation}
\item
  Compute the neutrino spectral temperature, assuming equilibrium with the
  fluid
  \begin{equation}
    T_\nu = \langle \varepsilon \rangle
    \frac{\mathscr{F}_2(\eta_\nu)}{\mathscr{F}_3(\eta_\nu)},
  \end{equation}
  where $\mathscr{F}_b$ is the Fermi integral
  \begin{equation}
    \label{eqn:fermi_integral}
    \mathscr{F}_b(\eta) = \int_0^\infty dx \, x^b (1+e^{x-\eta})^{-1}.
  \end{equation}
\item
  Assume the background neutrino fields have the same total density
  as the evolved moments
  \begin{align}
    \label{eqn:J_from_gray}
    J(\varepsilon) &=
    J \frac{\varepsilon^3}{T_\nu^4 \,\mathscr{F}_3(\eta_\nu)}
    (1+\exp(\varepsilon/T_\nu-\eta_\nu)^{-1}, \\
    \label{eqn:H_from_gray}
    H^\mu(\varepsilon) &=
    H^\mu \frac{\varepsilon^3}{T_\nu^4\, \mathscr{F}_3(\eta_\nu)}
    (1+\exp(\varepsilon/T_\nu-\eta_\nu)^{-1}.
  \end{align}
\end{enumerate}
When we use these moments in Eqn.~\ref{eqn:background_phi},
we call this the \emph{gray} method.

>>>>

\newcommand{\todo}[1]{\marginpar{\tiny{\textcolor{red}{#1}}}}
\renewcommand\todo[1]{} % if you wish to hide all todo notes

@@

\subsection{Post-Bounce Collapse Profile:
  Testing Scattering}
\label{ssec:test_collapse}
To test our scattering treatments we calculate neutrino fields outside a
collapsed 15~${\rm M}_\odot$ star, 100~ms after core bounce,
comparing ray tracing fields to those from a Monte Carlo transport calculation.
Elastic scattering interior to the shock at $r\approx150$~km
significantly modifies the neutrinos' spectra,
\todo{demonstrate}
and the extended envelope outside the shock becomes a source of
higher-than-average-energy neutrinos.

@@

The matter profile and background field are stored on a spherical
pseudo-spectral grid composed of 11 spherical-shell subdomains
\todo{cite early SpEC paper}
comprising a total of 62 radial grid points spaced approximately
logarithmically across $r\in(0,740)$~km.
The background scattering field $\Phi(\varepsilon)$ is supplied by
\lstinline{GR1D} in the form of $J(\varepsilon_i)$,
with $\varepsilon_i$ representing 18 energy groups
identical to those in the \lstinline{NuLib} table described in
Sec.~\ref{ssec:test_equilibrium}.
For the \emph{spectral} method we use $J(\varepsilon_i)$ directly,
using zeroth-order interpolation between energy groups;
for the \emph{gray} method we use $J=\sum J(\varepsilon_i) \Delta\varepsilon_i$,
and $G=\sum J(\varepsilon_i) \Delta\varepsilon_i/\varepsilon_i$ instead,
with $\Delta\varepsilon_i$ the bin-width of the $i$th energy group.

@@

As expected the average energies from both
the scattering envelope ($\cos A \lesssim 0.95$) and
the bright core ($\cos A\gtrsim 0.995 $)
are well characterized by the \emph{scat} treatments
and badly characterized by the \emph{noscat} treatment.
In this case for $\nu_x$ the major effect of elastic scattering
is to decrease the average energy of neutrinos coming from the core
and increase the average energy of neutrinos coming from the envelope.

@@

We can make some quantitative sense of the scattered energies in
Fig.~\ref{fig:avg_eps_collapse} and Tab.~\ref{tab:collapse_rt_vs_mc}
using the solutions explored in the simple configurations above.
In particular, the average energy in the scattering envelope
($\cos A\lesssim 0.95$) is related to the spectrum in the direction
of the interior ($\cos A \sim 1$).
The relation may be derived by simplifying our realistic model
to that of a homogeneous matter profile and background scattering field.

By expanding the exponential of
Eqns.~\ref{eqn:fse_homogeneous_general} and \ref{eqn:fae_homogeneous_general}
as we did for Eqn.~\ref{eqn:fae_homogeneous_limits},
and furthermore factoring out the dominant energy dependence from the
cross-sections (i.e.\ $\chi\equiv\varepsilon^2 \zeta$
with $\zeta$ approximately constant), we have
\begin{align}
  \label{eqn:fae_homogeneous_thin_limit}
  f_{\rm AE}(\varepsilon)
  &= \varepsilon^2 \zeta_a \,f^{\rm eq}(\varepsilon)\,s, \\
  \label{eqn:fse_homogeneous_thin_limit}
  f_{\rm SE}(\varepsilon)
  &= \varepsilon^2 \zeta_s \,\Phi(\varepsilon)\,s,
\end{align}
where $s$ is the length of scattering envelope passed through by the ray.
\todo{should be $\chi_a^*=\varepsilon\zeta_a$ for $\nu_x$}
In the envelope the local temperature is low so $f^{\rm eq}/\Phi\ll 1$,
and also $\zeta_a/\zeta_s\ll 1$,
so $f_{\rm SE}$ strongly dominates over $f_{\rm AE}$.
The average energy of the scattered field is therefore
\begin{equation}
  \label{eqn:avg_eps_sc}
  \langle \varepsilon \rangle^{\rm sc}
  = \frac{\int d\varepsilon\,\varepsilon^3 f}{\int d\varepsilon\,\varepsilon^2 f}
  \approx \frac{\int d\varepsilon\,\varepsilon^5 \Phi}{\int d\varepsilon\,\varepsilon^4 \Phi}.
\end{equation}

The spectrum of the background field $\Phi(\varepsilon)$ is well-approximated
as the \emph{scat} solution for $\cos A\sim 1$, since that is the dominant
source direction for neutrinos.
And because the scattering envelope is optically thin at all energies,
we can assume the neutrino spectrum is essentially unchanged in its passage
through the envelope.

In order to estimate $\langle \varepsilon \rangle^{\rm sc}$ analytically, we may
approximate the background field $\Phi(\varepsilon)$ analytically.
Direct Fermi-Dirac fits using a temperature and chemical potential
representative of a physical neutrinosurface fair poorly since neutrinos of
different energies decouple from the matter at different radii, over which
the thermodynamic state varies substantially.
Phenomenological Fermi-Dirac fits work well;
but so do pinched spectral fits which are much simpler \cite{miri????}:
\begin{equation}
  \label{eqn:pinched_spectra}
  \Phi^{\rm pi}(\varepsilon) \propto
  \varepsilon^{\alpha-2}
  \exp\left(-(\alpha+1)\frac{\varepsilon}{\langle\varepsilon\rangle}\right),
\end{equation}
(where our definition differs from that of \cite{miri????} by a factor of
$\varepsilon^2$, since we define our distribution function to be dimensionless
in natural units $\{\hbar,c\}=1$).

Pinching parameters
(calculated by eye from the \emph{spectral} \emph{scat} method)
for the species
$\{\nu_e,\bar{\nu}_e,\nu_x\}$ are $\alpha\approx\{3.6,5.1,2.3\}$.
Energy moments of pinched spectra like those in Eqn.~\ref{eqn:avg_eps_sc}
have a simple analytic form so that Eqn.~\ref{eqn:avg_eps_sc} becomes
\begin{align}
  \label{eqn:pinched_avg_eps_sc_1}
  \langle\varepsilon\rangle^{\rm sc}
  &\approx \frac{\alpha+3}{\alpha+1} \langle\varepsilon\rangle, \\
  %\label{eqn:pinched_avg_eps_sc_2}
  %&= \{1.4,1.3,1.6\} \langle\varepsilon\rangle, \\
  \label{eqn:pinched_avg_eps_sc_3}
  &\approx \{16,18,28\}\,~{\rm MeV},
\end{align}
again for $\{\nu_e,\bar{\nu}_e,\nu_x\}$ respectively.
These analytic predictions agree with the average energies of the scattering
envelope to approximately 10\%
% From by-eye examinations of <eps>(cosA) plot using data from
% MCDATA=$DIR/ResultsSherwood-02
% RTDATA_GR=$DIR/ResultsBrett-B-03/Test-scat-01/
% RTDATA_SP=$DIR/ResultsBrett-A-03/Test-scat-01/
%          nue     nua     nux
of all of the treatments including elastic scattering presented
in Tab.~\ref{tab:collapse_rt_vs_mc}, except for
$\bar{\nu}_e$ in the \emph{gray} \emph{scat} treatment,
which deviates from our prediction by 30\%.
This agreement is excellent despite the drastic
simplifications used in our toy model.

The $\langle\varepsilon_{\bar{\nu}_e}\rangle^{\rm sc}$ prediction
in the \emph{gray} \emph{scat} treatment is approximately 30\%
larger than the the fiducial Monte Carlo estimate. This is
due to the large negative local chemical potential
$\eta_{\bar{\nu}_e}\sim-10$ in the scattering envelope.
As described in Sec.~\ref{ssec:sources_se},
we use $\eta_\nu$ in the \emph{gray} treatment to construct our
synthetic spectrum from total neutrino densities $J$ and $G$.
This error points the way to future improvements to the \emph{gray} method.

@@

However moment evolutions with multiple energy groups are still rare.
Most simulations employ a gray moment scheme,
evolving only the energy-integrated moments, $J$ and $H^\alpha$
and sometimes the number density $G$.
If such is the case we resort to the \emph{gray} method,
by reconstructing the energy-dependent source terms
from gray moments, assuming a diluted Fermi-Dirac spectrum.
We use the following procedure:
\begin{enumerate}
\item
  Interpolate the fluid temperature, $T$,
  equilibrium neutrino chemical potential, $\eta_\nu$,
  and evolved neutrino energy and number densities in the fluid frame,
  $J$ and $G$, from the simulation grid.
\item
  Compute the average neutrino energy in the fluid frame
  \begin{equation}
    \langle \varepsilon \rangle \equiv J/G.
  \end{equation}
\item
  Compute the neutrino spectral temperature, assuming equilibrium with the
  fluid
  \begin{equation}
    \label{eqn:spectral_temp}
    T_\nu = \langle \varepsilon \rangle
    \frac{\mathscr{F}_2(\eta_\nu)}{\mathscr{F}_3(\eta_\nu)},
  \end{equation}
  where $\mathscr{F}_b$ is the Fermi integral
  \begin{equation}
    \label{eqn:fermi_integral}
    \mathscr{F}_b(\eta) = \int_0^\infty dx \, x^b (1+e^{x-\eta})^{-1}.
  \end{equation}
\item
  Assume the background neutrino fields have the same total density
  as the evolved moments
  \begin{align}
    \label{eqn:J_from_gray}
    J(\varepsilon) &=
    J \frac{\varepsilon^3}{T_\nu^4 \,\mathscr{F}_3(\eta_\nu)}
    (1+\exp(\varepsilon/T_\nu-\eta_\nu)^{-1}, \\
    \label{eqn:H_from_gray}
    H^\mu(\varepsilon) &=
    H^\mu \frac{\varepsilon^3}{T_\nu^4\, \mathscr{F}_3(\eta_\nu)}
    (1+\exp(\varepsilon/T_\nu-\eta_\nu)^{-1}.
  \end{align}
\end{enumerate}
When we use these moments in Eqn.~\ref{eqn:background_phi},
we call this the \emph{gray} method.
